---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Setup the directory structure, filenames etc:
```{r warning=FALSE}
data.dir <- "data"
dir.create(data.dir, recursive=TRUE)
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zipfile <- "activity.zip"
datafile <- "activity.csv"
```

Obtain the raw data file by doing the minimum work to get it. 
```{r warning=FALSE}
setwd(data.dir)
if (!file.exists(datafile)) {
  if (!file.exists(zipfile)) {
    download.file(url=url, destfile=zipfile, method="wget")
  }
  unzip(zipfile)
}
datafile <- paste0(data.dir, "/", datafile)
```

Read in the csv file, convert to a tibble (I'm a 'hadleyverse' fan!) and clean it of any NAs
```{r echo=FALSE, warning=FALSE}
require(dplyr)
raw.data <- tbl_df(read.csv(datafile))
clean.data <- na.omit(raw.data)
```

## What is mean total number of steps taken per day?
```{r results=asis}
x <- group_by(clean.data, date) %>% summarize(total_steps_per_day=sum(steps)) %>% summarize(mean_total_steps_per_day=mean(total_steps_per_day))
x$mean_total_steps_per_day
```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
